---
title: "Household Power consumption - Time Series Analysis"
output: html_notebook
---

```{r setup, warning=F, message=F,echo=F}
library(tibble)
library(dplyr)
library(tidyr)
library(readr)
library(lubridate)
library(ggplot2)
library(tsibble)
library(fable)
library(fabletools)
library(feasts)
library(tsibbledata)
library(lubridate)
library(forecast)
library(zoo)
library(tseries)
library(corrplot)
```

1.date: Date in format dd/mm/yyyy
2.time: time in format hh:mm:ss
3.global_active_power: household global minute-averaged active power (in kilowatt)
4.global_reactive_power: household global minute-averaged reactive power (in kilowatt)
5.voltage: minute-averaged voltage (in volt)
6.global_intensity: household global minute-averaged current intensity (in ampere)
7.sub_metering_1: energy sub-metering No. 1 (in watt-hour of active energy). It corresponds to the kitchen, containing mainly a dishwasher, an oven and a microwave (hot plates are not electric but gas powered).
8.sub_metering_2: energy sub-metering No. 2 (in watt-hour of active energy). It corresponds to the laundry room, containing a washing-machine, a tumble-drier, a refrigerator and a light.
9.sub_metering_3: energy sub-metering No. 3 (in watt-hour of active energy). It corresponds to an electric water-heater and an air-conditioner.

```{r}

power_data <- read.table("household_power_consumption.txt", sep = ';', header = TRUE, na.strings = "?") %>%
  mutate(datetime = dmy_hms(paste(Date, Time))) %>%
  select(datetime, Global_active_power, Global_reactive_power, Voltage, Global_intensity, Sub_metering_1, Sub_metering_2, Sub_metering_3) %>%
  tsibble(index = datetime)

head(power_data)

```
```{r}
power_data_filled <- power_data %>%
  mutate(across(where(is.numeric), na.interp))

head(power_data_filled)
```
Interpolation was used to fill the empty gaps in data, as methods like back-filling would not suffice for the length of gaps where the data was missing. 

```{r}
ggplot(power_data_filled[0:60000, ], aes(x = datetime, y = Global_active_power)) +
  geom_line() +
  labs(
    title = "Time Series plot of Global active power in the household", 
    x = "Time", 
    y = "Global active power"
  ) +
  theme_minimal()

```

```{r}
monthly_ga_power <- power_data_filled %>%
  index_by(month = ~ yearmonth(.)) %>%
  summarize(
    average_ga_power = mean(Global_active_power, na.rm = TRUE),
    total_ga_power = sum(Global_active_power, na.rm = TRUE)
  )

head(monthly_ga_power)
```

```{r}

ggplot(monthly_ga_power, aes(x = month, y = average_ga_power)) +
  geom_line() +
  labs(
    title = "Time Series plot of monthly average Global active power",
    x = "Monthly Average",
    y = "Month"
  )

```
```{r}

ggplot(monthly_ga_power, aes(x = month, y = total_ga_power)) +
  geom_line() +
  labs(
    title = "Time Series plot of monthly total Global active power",
    x = "Monthly Total",
    y = "Month"
  )

```



```{r}

monthly_ga_power %>%
  model(
    classical_decomposition(average_ga_power, type = "multiplicative")
  ) %>%
  components() %>%
  autoplot()

```

There is a seasonal element to the monthly average global active power for the household, with a peak in average usage during the months of Jan. 

Analysing seasonal trends for the 3 submeter readings... 


```{r}

monthly_sm_power <- power_data_filled %>%
  index_by(month = ~ yearmonth(.)) %>%
  summarize(
    sm1_avg = mean(Sub_metering_1, na.rm = TRUE),
    sm1_total = sum(Sub_metering_1, na.rm = TRUE),
    sm2_avg = mean(Sub_metering_2, na.rm = TRUE),
    sm2_total = sum(Sub_metering_2, na.rm = TRUE),
    sm3_avg = mean(Sub_metering_3, na.rm = TRUE),
    sm3_total = sum(Sub_metering_3, na.rm = TRUE)
  )

head(monthly_sm_power)

```

```{r}
ggplot(monthly_sm_power, aes(x = month, y = sm1_avg)) +
  geom_line() +
  labs(
    title = "Time Series plot of monthly Average reading for submeter 1",
    x = "Monthly Avg",
    y = "Month"
  )

```

```{r}
ggplot(monthly_sm_power, aes(x = month, y = sm2_avg)) +
  geom_line() +
  labs(
    title = "Time Series plot of monthly Average reading for submeter 2",
    x = "Monthly Avg",
    y = "Month"
  )

```

```{r}
ggplot(monthly_sm_power, aes(x = month, y = sm3_avg)) +
  geom_line() +
  labs(
    title = "Time Series plot of monthly Average reading for submeter 3",
    x = "Monthly Avg",
    y = "Month"
  )

```
```{r}

monthly_sm_power %>%
  model(
    classical_decomposition(sm1_avg, type = "multiplicative")
  ) %>%
  components() %>%
  autoplot()

```
```{r}

monthly_sm_power %>%
  model(
    classical_decomposition(sm2_avg, type = "multiplicative")
  ) %>%
  components() %>%
  autoplot()

```

```{r}

monthly_sm_power %>%
  model(
    classical_decomposition(sm3_avg, type = "multiplicative")
  ) %>%
  components() %>%
  autoplot()

```

All 3 submeter readings have a dip during summer months and a peak during winter months. This is to be expected for submeter 3, as it corresponds to the electric water-heater and air conditioner, but it is interesting to observe that the other 2 submeter readings follow a similar seasonality. 

ETS modelling for global active power:

```{r}
fit_ets_1 <- monthly_ga_power %>%
  model(ETS(average_ga_power ~ error("M") + trend("M") + season("M")))

report(fit_ets_1)
```
```{r}
fit_ets_2 <- monthly_ga_power %>%
  model(ETS(log(average_ga_power) ~ error("M") + trend("M") + season("M")))

report(fit_ets_2)

```
```{r}

fit_ets_3 <- monthly_ga_power %>%
  model(ETS(average_ga_power ~ error("A") + trend("A") + season("A")))

report(fit_ets_3)

```

```{r}

gg_tsresiduals(fit_ets_3)

```
```{r}
cutoff <- yearmonth(as.Date(max(monthly_ga_power$month)) - months(5))

monthly_ga_power_train <- monthly_ga_power %>%
  filter(month < cutoff)
monthly_ga_power_test <- monthly_ga_power %>%
  filter(month >= cutoff)

tail(monthly_ga_power_train)

fit_ets_test <- monthly_ga_power_train %>%
  model(ETS(average_ga_power ~ error("A") + trend("A") + season("A")))

fc_ets <- forecast(fit_ets_test, h=6) 
fc_ets %>% autoplot(monthly_ga_power_test)
```
```{r}
print(accuracy(fc_ets, monthly_ga_power_test))
```

```{r}
fit_snaive <- monthly_ga_power_train %>%
  model(SNAIVE(average_ga_power))
gg_tsresiduals(fit_snaive)
```
```{r}
report(fit_snaive)

```

```{r}

fc_snaive <- forecast(fit_snaive, h=6) 
fc_snaive %>% autoplot(monthly_ga_power_test)

```
```{r}

print(accuracy(fc_snaive, monthly_ga_power_test))

```
```{r}
monthly_summary_all <- power_data_filled %>%
  index_by(month = ~ yearmonth(.)) %>%
  summarize(
    average_ga_power = mean(Global_active_power, na.rm = TRUE),
    average_gr_power = mean(Global_reactive_power, na.rm = TRUE),
    average_voltage = mean(Voltage, na.rm = TRUE),
    average_gi = mean(Global_intensity, na.rm = TRUE),
    average_sm1 = mean(Sub_metering_1, na.rm = TRUE),
    average_sm2 = mean(Sub_metering_2, na.rm = TRUE),
    average_sm3 = mean(Sub_metering_3, na.rm = TRUE)
  )

head(monthly_summary_all)

```

```{r}

corr_matrix <- as.data.frame(monthly_summary_all) %>%
  select(-month) %>%
  cor()

print(corr_matrix)

```
Global active power has a string relationship between global intensity and the 3 submeter readings, hence we will use those in a multiple regression forecasting model
```{r}
monthly_summary_train <- monthly_summary_all %>%
  filter(month < cutoff)
monthly_summary_test <- monthly_summary_all %>%
  filter(month >= cutoff)
head(monthly_summary_train)
```

```{r}

fit_lm <- lm(average_ga_power ~ average_gi + average_sm1 + average_sm2 + average_sm3 + month, data = monthly_summary_train)

summary(fit_lm)

```

```{r}
fc_lm <- predict(fit_lm, monthly_summary_test)

monthly_summary_test$predicted <- fc_lm

ggplot() +
  geom_line(data=monthly_summary_test, aes(x=month, y=average_ga_power, color="Test Data Actual")) +
  geom_line(data=monthly_summary_test, aes(x=month, y=predicted, color="Test Data Predicted")) +
  labs(title = "Forecast using multiple regression model", 
       x = "Month", y = "Average Global Active Power") +
  scale_color_manual(values = c("Train Data" = "blue", 
                               "Test Data Actual" = "red", 
                               "Test Data Predicted" = "green")) +
  theme_minimal()
```
```{r}
ggplot(monthly_ga_power_train, aes(x = month, y = residuals(fit_lm))) +
  geom_line(color = "blue") +
  labs(title = "Residuals of the Multiple Regression Model",
       x = "Month", y = "Residuals") +
  theme_minimal()

```
Checks for ARIMA

```{r}
lambda <- BoxCox.lambda(monthly_ga_power$average_ga_power)
transformed_data <- BoxCox(monthly_ga_power$average_ga_power, lambda)

monthly_ga_power <- monthly_ga_power %>%
  mutate(transformed_avg_ga_power = transformed_data)

# Plot the transformed data
ggplot(monthly_ga_power, aes(x = month, y = transformed_avg_ga_power)) +
  geom_line(color = "purple") +
  labs(title = "Box-Cox Transformed Data", x = "Time", y = "Transformed Power") +
  theme_minimal()

```
Stationarity Test and Differencing

test the stationarity of the transformed data using the Augmented Dickey-Fuller (ADF) test. If the data is non-stationary (p-value > 0.05), apply differencing to make it stationary.

```{r}
# Perform ADF test
adf_test <- adf.test(monthly_ga_power$transformed_avg_ga_power, alternative = "stationary")
print(adf_test)

# Differencing if necessary
if (adf_test$p.value > 0.05) {
  monthly_ga_power <- monthly_ga_power %>%
    mutate(diff_transformed_avg_ga_power = difference(transformed_avg_ga_power))
}

# Plot the data after differencing (if applicable)
ggplot(monthly_ga_power, aes(x = month, y = transformed_avg_ga_power)) +
  geom_line(color = "blue") +
  labs(title = "Transformed Data (Stationary)", x = "Time", y = "Transformed Average Power") +
  theme_minimal()

```
Examine ACF and PACF
The Autocorrelation Function (ACF) and Partial Autocorrelation Function (PACF) plots help in identifying the order of AR (p) and MA (q) components of the ARIMA model.

```{r}
# ACF and PACF plots
ggAcf(monthly_ga_power$transformed_avg_ga_power) +
  labs(title = "ACF of Transformed Data")

ggPacf(monthly_ga_power$transformed_avg_ga_power) +
  labs(title = "PACF of Transformed Data")

```
Fitting ARIMA model
```{r}

monthly_ga_power_train <- monthly_ga_power %>%
  filter(month < cutoff)
monthly_ga_power_test <- monthly_ga_power %>%
  filter(month >= cutoff)

fit_arima <- monthly_ga_power_train %>%
  model(ARIMA(transformed_avg_ga_power))

report(fit_arima)

```
```{r}

gg_tsresiduals(fit_arima)

```

```{r}
fc_arima <- fit_arima %>%
  forecast(h=6)

fc_arima %>% autoplot(monthly_ga_power_test)
```
```{r}
print(accuracy(fc_arima, monthly_ga_power_test))
```

```{r}

forecast_summary <- data.frame(month = monthly_summary_test$month,
                               actual = monthly_summary_test$average_ga_power,
                               snaive = fc_snaive$.mean,
                               ets = fc_ets$.mean,
                               arima = fc_arima$.mean,
                               mlr = fc_lm)
head(forecast_summary)
```

```{r}

ggplot(monthly_summary_all, aes(x=month)) + 
  geom_line(data = monthly_summary_all, aes(y=average_ga_power, color = "Actual")) +
  geom_line(data = monthly_ga_power, aes(y=transformed_avg_ga_power, color = "Transformed")) +
  geom_line(data = forecast_summary, aes(y = snaive, color = "SNAIVE")) +
  geom_line(data = forecast_summary, aes(y = arima, color = "ARIMA")) +
  geom_line(data = forecast_summary, aes(y = ets, color = "ETS")) +
  geom_line(data = forecast_summary, aes(y = mlr, color = "MLR")) +
  labs(title = "Forecast Summary",
       x = "Month", y = "Average Global Active Power") +
  scale_color_manual(values = c("Actual" = "black","Transformed" = "navy", "SNAIVE" = "blue",
                                "ARIMA" = "skyblue","ETS" = "red", "MLR" = "green")) +
  theme_minimal() +
  theme(legend.position = "top")

```



